---
title: "National BioBlitz 2025 – League Tables"
author: "The Rock Pool Project"
output: html_document
fontsize: 10pt
mainfont: Montserrat
sansfont: Montserrat
monofont: Courier New
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## League Tables – National BioBlitz 2025

This report showcases three league tables based on iNaturalist data from the National BioBlitz 2025. Each table highlights different aspects of participant contributions and biodiversity findings, filtered to include only high-quality, research-grade records.

All data have been cleaned, filtered, and scored according to the methods outlined in earlier training sessions.

## 1. League Table: Research-Grade Records

This table ranks participants based solely on the number of **research-grade observations** they submitted. Research-grade records meet iNaturalist's standard for community verification, meaning they have:

 - A valid species-level ID,
 - A date, time, and location,
 - At least two confirming identifications.

This provides a simple but reliable measure of participation and data quality, which is particularly important when a competition is involved.


```{r data}
#load the latest data
saved_data_path <- "../NatBioBlitz_iNat.RData"

if (file.exists(saved_data_path)) {
  load(saved_data_path)
} else {
  source("scripts/new get project obs function.R")
  NatBioBlitz_iNat <- get_inat_obs_project_v2("brpc-national-bioblitz-2025")
  save(NatBioBlitz_iNat, file = saved_data_path)
}
```

```{r league1}
#subset to research grade and species level observations only:

NatBioBlitz_iNat_RG <- subset(NatBioBlitz_iNat, quality_grade == "research" & taxon.rank == "species")

#create league
library(tidyverse)

league_table1 <- NatBioBlitz_iNat_RG %>%
  group_by(user.login) %>%
  summarise(
    total_records = n(),
    total_species = n_distinct(taxon.name[taxon.rank == "species"])
  ) %>%
  arrange(desc(total_species))

#create names vector
all_users <- subset(NatBioBlitz_iNat, !duplicated(NatBioBlitz_iNat$user.login))

all_users_names <- all_users$user.name
names(all_users_names) <- all_users$user.login

all_users_names[is.na(all_users_names) | all_users_names == ""] <- all_users$user.login[is.na(all_users_names) | all_users_names == "" ]

# add to data table
NatBioBlitz_iNat$user <- all_users_names[NatBioBlitz_iNat$user.login]


#use these names in the league table

league_table1$user.login <- all_users_names[league_table1$user.login]

#fix column heading 1

names(league_table1)[1] <- "Observer"

library(kableExtra)

kable(league_table1, caption = "National BioBlitz League Table – Research Grade only")

```

## 2. League Table: Total Species Rarity Score

In this version of the league table, each species has been assigned a **rarity-based score**. Rarer species contribute more points to a participant’s total, encouraging both participation and biodiversity exploration.

Scores have been applied only to **species-level, research-grade records**. The total score for each participant reflects the sum of all scored observations they contributed.

This approach helps to highlight users who observed uncommon or under-recorded marine life.


```{r species_scores}
#get species scores
load("../Data/Latest_UK_Species_Scores.RData")

sp_scores <- UK_N_obs$Score
names(sp_scores) <- as.character(UK_N_obs$id)

#add scores to iNat data
NatBioBlitz_iNat$Score <- sp_scores[as.character(NatBioBlitz_iNat$taxon.id)]

#if species level observation not have a score, give is zero unless it's Research grade then give it a 20
NatBioBlitz_iNat$Score[
  is.na(NatBioBlitz_iNat$Score) &
  NatBioBlitz_iNat$taxon.rank == "species" &
  NatBioBlitz_iNat$quality_grade == "research"
] <- 20

# double the scores for the non-native species
# Load the non-native species list

non_native_species <- read.csv("../Data/UK marine NNS.csv")

NatBioBlitz_iNat$Score[NatBioBlitz_iNat$taxon.id %in% non_native_species$inat_id] <- NatBioBlitz_iNat$Score[NatBioBlitz_iNat$taxon.id %in% non_native_species$inat_id] * 2

#recreate RG subset
NatBioBlitz_iNat_RG <- subset(NatBioBlitz_iNat, quality_grade == "research" & taxon.rank == "species")


league_table2 <- NatBioBlitz_iNat_RG %>%
  group_by(user.login) %>%
  summarise(
    total_records = n(),
    total_species = n_distinct(taxon.name[taxon.rank == "species"]),
    total_score = sum(Score, na.rm = TRUE)
  ) %>%
  arrange(desc(total_score))

#use real names in the league table where available
league_table2$user.login <- all_users_names[league_table2$user.login]

#fix column heading 1

names(league_table2)[1] <- "Observer"


kable(league_table2, caption = "National BioBlitz League Table – Rarity Scores")


```

## 3. League Table: Top Species per Observer

This table shows each observer's **single highest-scoring species**. It celebrates individual record highlights and lets us explore which rare or notable species were captured by whom.

It’s a useful way to recognise standout contributions, especially when observers find something unique—even if they only submitted a few records overall.

```{r league3}
league_table3 <- NatBioBlitz_iNat_RG %>%
  group_by(user.login) %>%
  summarise(
    max_score = max(Score, na.rm = TRUE)
  ) %>%
  arrange(desc(max_score))

league_table3$Common <- NA
league_table3$Scientific <- NA


for (observer in league_table3$user.login) {
  obs_dat <- subset(NatBioBlitz_iNat_RG, user.login == observer)
  obs_best_sp <- subset(obs_dat, Score == max(Score))
  
  #just in case there are more than one
  obs_best_sp <- obs_best_sp[sample(1:nrow(obs_best_sp),1),]
  
  #add to table
  league_table3$Common[league_table3$user.login == observer] <- obs_best_sp$taxon.preferred_common_name
  
  league_table3$Scientific[league_table3$user.login == observer] <- obs_best_sp$taxon.name
}

#use real names in the league table where available
league_table3$user.login <- all_users_names[league_table3$user.login]

#fix column heading 1

names(league_table3)[1] <- "Observer"


kable(league_table3, caption = "National BioBlitz League Table – Max Rarity Score")


```
