---
output:
  pdf_document:
    latex_engine: xelatex
    includes:
      in_header: report-header.tex
fontsize: 10pt
mainfont: Montserrat
sansfont: Chivo
monofont: Courier New
geometry: a4paper, top=30mm, bottom=20mm, left=20mm, right=20mm
urlcolor: blue
---

\begin{center}

{\sffamily\fontsize{28pt}{34pt}\selectfont \textbf{Big Rock Pool Challenge}}

\end{center}

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)

library(ggplot2)
library(dplyr)
library(showtext)
library(ggtext)
library(ggimage)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(knitr)

# Fonts for plots
font_add_google("Montserrat", "mont")
font_add_google("Chivo", "chivo")
showtext_auto()
```

```{r data_load}
load("NatBioBlitz_iNat20250604.RData") #iNat BioBlitz Data as of 4th June 2025

```

\begin{center}
\includegraphics[width=0.8\textwidth]{images/natbioadvert.png}

{\sffamily\fontsize{24pt}{30pt}\selectfont \textbf{BioBlitz Report}}

\includegraphics[width=0.8\textwidth]{images/BRPC Funders and Partner logos.png}
\end{center}

\newpage



## Executive Summary

A brief overview of the National BioBlitz 2025: aims, reach, and key achievements. Include highlights and a general snapshot of participation.

\newpage



# About the National BioBlitz

The [National BioBlitz 2025](https://www.therockpoolproject.co.uk/national-bioblitz) was a UK-wide citizen science initiative coordinated by The Rock Pool Project, aimed at engaging people across the country in discovering and recording the biodiversity of our rocky shores.

Held between 17th and 25th May 2025, the event empowered individuals, families, and wildlife groups to take part in a shared national effort to explore intertidal habitats and contribute meaningful data to marine research and conservation.

Participants of all ages and backgrounds visited their local coastlines to record the species they found in rock pools and along the shore. Whether seasoned naturalists or first-time explorers, everyone played a role in helping to build a clearer picture of the UK's marine life.

## Data collection

All observations were recorded using [iNaturalist](https://www.inaturalist.org/), a free, community-powered biodiversity recording platform. Participants submitted photographs of their finds, along with the date and location of each observation. These were automatically grouped into a dedicated project: the [National BioBlitz 2025 iNaturalist page](https://www.inaturalist.org/projects/brpc-national-bioblitz-2025), which captured all qualifying submissions made during the event period.

Species identifications were suggested by participants and verified through the iNaturalist community and The Rock Pool Project’s team of marine biologists. This process helped ensure data quality while also offering an accessible learning experience for participants.

The resulting data set represents a unique, community-generated snapshot of intertidal biodiversity across the UK — supporting long-term research and helping to track changes in species distributions, including the spread of marine non-natives.



# Results
The 2025 National BioBlitz brought together an inspiring community of nature enthusiasts, families, schools, and volunteers to explore the UK’s rocky shores. 

```{r overall-summary-plot1}
# Infographic One - data summary

## first save the latest summary data
## How many records?
n_recs <- nrow(NatBioBlitz_iNat)

## How many observers?
observers <- unique(NatBioBlitz_iNat$user.login)

n_observers <- length(observers)

#how many species

sp_dat <- subset(NatBioBlitz_iNat, taxon.rank == "species")

sp_list <- unique(sp_dat$taxon.name)

n_sp <- length(sp_list)

# Load required libraries
library(ggplot2)    # Core plotting library
library(ggtext)     # Allows rich text (e.g. markdown) in plot titles and labels
library(ggimage)    # Enables image placement inside ggplots
library(showtext)   # Allows use of custom fonts (e.g. Google Fonts)

# Load the RPP Google Font "Montserrat" and name it "mont" in R
font_add_google("Montserrat", "mont")

# Activate showtext so custom fonts are rendered in plots
showtext_auto()

# Create the dataset with three summary metrics and corresponding icons
summary_data <- data.frame(
  label = c("observers", "records", "Species"),  # Labels for the metrics
  value = c(n_observers, n_recs, n_sp),                       # Actual numbers to display
  icon = c("icons/people.png",                  # Path to PNG icons (must exist)
           "icons/clipboard.png",
           "icons/crab.png")
)

# Create formatted labels using HTML-style markup
summary_data$label_combined <- paste0(
  "<b>", format(summary_data$value, big.mark = ","), "<br>",
  summary_data$label, "</b>"
)

plot1 <- ggplot(summary_data, aes(x = label, y = 1)) +
  geom_image(aes(image = icon, y = 0.82), size = 0.4) +

  geom_richtext(
    aes(y = 0.63, label = label_combined),
    family = "mont",
    fill = NA, label.color = NA,  # Transparent background
    size = 10,
    lineheight = 1,
    hjust = 0.5
  ) +

  coord_cartesian(ylim = c(0.55, 0.9)) +
  theme_void(base_family = "mont") +
  theme(
    plot.margin = margin(5, 10, 5, 10)
  )


# Draw the plot to the file
plot1

```

These headline figures highlight both the breadth of public engagement and the diversity of marine life encountered across the country’s intertidal zones. Each record contributes to a growing national dataset that supports long-term marine monitoring, biodiversity research, and the detection of non-native species. 

## Geographical Coverage

```{r geo_coverage}
# Convert 'location' column into latitude and longitude
# ------------------------------
# 'location' field is a string like "50.12,-5.06" – we split it into two numbers
location_split <- strsplit(NatBioBlitz_iNat$location, ",")
location_df <- do.call(rbind, lapply(location_split, function(x) as.numeric(x)))
colnames(location_df) <- c("latitude", "longitude")  # Label the columns

# Add these new lat/lon values to the main dataset
NatBioBlitz_iNat <- cbind(NatBioBlitz_iNat, location_df)

# Remove any rows that failed to convert properly
NatBioBlitz_iNat <- NatBioBlitz_iNat %>%
  filter(!is.na(latitude), !is.na(longitude))

# Convert the data frame to a spatial object
obs_points <- st_as_sf(NatBioBlitz_iNat, coords = c("longitude", "latitude"), crs = 4326)

# ------------------------------
# Load a background map of the UK and Ireland
# ------------------------------
uk_map <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin %in% c("United Kingdom", "Ireland"))  # Keep only UK and Ireland

# ------------------------------
# Create the map plot
# ------------------------------
plot2 <- ggplot() +
  geom_sf(data = uk_map, fill = "whitesmoke", colour = "grey50") +       # Draw base map
  geom_sf(data = obs_points, colour = "#0B6EF5", size = 5, alpha = 0.7) + # Plot observation points
  coord_sf(xlim = c(-11, 3), ylim = c(48.5, 61), expand = FALSE) +       # Zoom to UK
  labs(title = "Locations of National BioBlitz Observations (2025)") +   # Plot title
  theme_void(base_family = "mont") +                                     # Minimal theme with font
  theme(
    plot.title = element_text(
      family = "chivo", size = 16, hjust = 0.5,
      margin = margin(t = 10, b = 10)
    ),
    plot.margin = margin(10, 10, 10, 10)
  )

plot2

```
The map above shows the wide distribution of observations submitted during the 2025 National BioBlitz. From the remote shores of Shetland to the rocky coasts of Cornwall, participants across the UK came together to record marine life from their local intertidal habitats.

This nationwide coverage reflects the strength of our community-led approach — with contributions from both established hubs and brand-new participants, including schools, wildlife groups, and families exploring their nearby coastline.

The broad spread of data not only captures regional variation in species distributions, but also enhances the national value of the dataset — helping us build a clearer picture of the UK's marine biodiversity in a time of environmental change.

## Taxonomic Coverage

```{r taxo_coverage}
# Load the lookup table for iconic taxa
RPP_iconic_taxa <- read.csv("BRPC_iconic_taxa.csv", stringsAsFactors = FALSE)
RPP_iconic_taxa$iNat_code <- as.character(RPP_iconic_taxa$iNat_code)

# Create a named vector to map ancestry codes to taxon names
taxon_lookup <- setNames(RPP_iconic_taxa$Common, RPP_iconic_taxa$iNat_code)

# Define a function to search each ancestry string and return the first matching label
get_iconic_label <- function(ancestry_str) {
  ids <- unlist(strsplit(ancestry_str, "/"))
  match <- taxon_lookup[ids]
  label <- match[!is.na(match)][1]
  if (is.na(label)) "Other" else label
}

# Apply the function to all records
NatBioBlitz_iNat$iconic_taxon <- vapply(NatBioBlitz_iNat$taxon.ancestry, get_iconic_label, character(1))

# ------------------------------
# Count records by iconic taxon
# ------------------------------
iconic_counts <- NatBioBlitz_iNat %>%
  count(iconic_taxon, name = "n") %>%
  arrange(desc(n))

# Updated brand colour palette including algae groups
brand_colours <- c(
  "Crustaceans" = "#535C7D",     # Soft Grey
  "Red seaweed"    = "#DA3737",    # 
  "Green seaweed"  = "#D1F4FF",    # Crest Foam
  "Brown seaweed"  = "#FDBD19",    # Golden Shore
  "Molluscs"     = "#00A6FB",    # Ocean Spark
  "Fishes"         = "#4D56F5",    # Deep Blue (additional)
  "Sea anemones"   = "#FF9895",    # Coral tone
  "Echinoderms"  = "#FFDDA3",    # Pale Orange
  "Sponges"      = "#A3C4F3",    # Soft Blue
  "Worms"        = "#FFC0BF", #Soft Rose Shell
  "Sea squirts"  = "#2176FF",    # Tidal Blue
  "Other"        = "#191D2D"     # Midnight Tide (neutral)
)

# Ensure factor order matches
iconic_counts$iconic_taxon <- factor(iconic_counts$iconic_taxon, levels = names(brand_colours))


# ------------------------------
# Create the bar plot
# ------------------------------
plot3 <- ggplot(iconic_counts, aes(x = reorder(iconic_taxon, n), y = n, fill = iconic_taxon)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "National BioBlitz Records by Rock Pool Iconic Taxa",
    x = NULL,
    y = "Number of Records"
  ) +
  scale_fill_manual(values = brand_colours) +
  theme_minimal(base_family = "mont") +
  theme(
    plot.title = element_text(family = "chivo", size = 16, face = "bold", hjust = 0.5, margin = margin(b = 15)),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    plot.margin = margin(t = 20, r = 30, b = 30, l = 30)
  )

plot3

```

\newpage

## Best finds


```{r best_finds}
#add rarity scores
load("../Data/Latest_UK_Species_Scores.RData")

sp_scores <- UK_N_obs$Score
names(sp_scores) <- as.character(UK_N_obs$id)

#add scores to iNat data
NatBioBlitz_iNat$Score <- sp_scores[as.character(NatBioBlitz_iNat$taxon.id)]

#if species level observation not have a score, give is zero unless it's Research grade then give it a 20
NatBioBlitz_iNat$Score[
  is.na(NatBioBlitz_iNat$Score) &
  NatBioBlitz_iNat$taxon.rank == "species" &
  NatBioBlitz_iNat$quality_grade == "research"
] <- 20

# double the scores for the non-native species
# Load the non-native species list

non_native_species <- read.csv("../Data/UK marine NNS.csv")

NatBioBlitz_iNat$Score[NatBioBlitz_iNat$taxon.id %in% non_native_species$inat_id] <- NatBioBlitz_iNat$Score[NatBioBlitz_iNat$taxon.id %in% non_native_species$inat_id] * 2


NatBioBlitz_iNat$NNS <- NatBioBlitz_iNat$taxon.id %in% non_native_species$inat_id

#best native finds
NatBioBlitz_iNat_N <- subset(NatBioBlitz_iNat, !NNS & quality_grade == "research")
NatBioBlitz_iNat_N <- NatBioBlitz_iNat_N[order(NatBioBlitz_iNat_N$Score, decreasing = T),]

NatBioBlitz_iNat_N_best <- subset(NatBioBlitz_iNat_N, Score == max(Score, na.rm = T))

#best non-native finds
NatBioBlitz_iNat_NNS <- subset(NatBioBlitz_iNat, NNS & quality_grade == "research")
NatBioBlitz_iNat_NNS <- NatBioBlitz_iNat_NNS[order(NatBioBlitz_iNat_NNS$Score, decreasing = T),]

NatBioBlitz_iNat_NNS_best <- subset(NatBioBlitz_iNat_NNS, Score >= 10)


```


### Native species

We found some amazing species. These were the highest ranked according to our rarity points:

```{r best_native_recs, results='asis'}
#get photos
dir.create("images/best_native", showWarnings = FALSE)

for (i in 1:nrow(NatBioBlitz_iNat_N_best)) {
  
  url <- NatBioBlitz_iNat_N_best$observation_photos[[i]][1,]$photo.url
  
  #switch to medium image
  url <- gsub("square", "medium", url)
  
  if(!NatBioBlitz_iNat_N_best$license_code[[i]] %in% c("", "All Rights Reserved") | is.na(NatBioBlitz_iNat_N_best$license_code[[i]])){
  
  if (!is.na(url) && nzchar(url)) {
    filename <- paste0("images/best_native/native_", i, ".jpg")
    
    if(!filename %in% list.files("images/best_native")){
    download.file(url, destfile = filename, mode = "wb")
    NatBioBlitz_iNat_N_best$local_photo[i] <- filename  
    }
    
  } else {
    NatBioBlitz_iNat_N_best$local_photo[i] <- NA
  }
  }else{
    NatBioBlitz_iNat_N_best$local_photo[i] <- "License not available"
  }
}


# Loop through and print markdown for each record
for (i in 1:nrow(NatBioBlitz_iNat_N_best)) {
  rec <- NatBioBlitz_iNat_N_best[i, ]
  
  cat("\n\n")
  
  # Start minipage block (centered)
  cat("\\begin{center}\n")
  cat("\\begin{minipage}{0.9\\textwidth}\n")
  
  # Add image (sized down for PDF)
  if (rec$local_photo != "") {
    cat(sprintf("\\begin{center}\\includegraphics[height=6cm]{%s}\\end{center}\n\n", rec$local_photo))
  }
  
  # Print species and observation details
  if(!is.na(rec$taxon.preferred_common_name)){
  cat(sprintf("\\textbf{%s} \\textit{(%s)}\\\\\n", 
              rec$taxon.preferred_common_name, rec$taxon.name))
  }else{
    cat(sprintf("\\textbf{\\textit{%s}}\\\\\n", rec$taxon.name))
  }
  cat(sprintf("\\textbf{Rarity points:} %s\\\\\n", rec$Score))
  cat(sprintf("\\textbf{Location:} %s\\\\\n", rec$place_guess))
  
  if(!is.na(rec$user.name)){
    cat(sprintf("\\textbf{Observer:} %s\\\\\n", rec$user.name))
  }else{
    cat(sprintf("\\textbf{Observer:} %s\\\\\n", rec$user.login))
  }
  
  
  cat(sprintf("\\href{https://www.inaturalist.org/observations/%s}{View on iNaturalist}\\\\\n", rec$id))
  
    # Attribution if applicable
  if (!is.na(rec$observation_photos[[1]]$photo.attribution[1])) {
    cat(sprintf("\\textit{Photo licensed under CC-%s}\\\\\n", rec$observation_photos[[1]]$photo.attribution[1]))
  }

  # End minipage and add spacing
  cat("\\end{minipage}\n")
  cat("\\end{center}\n")
  cat("\\vspace{10mm}\n\n")
}

```
All of these records received a maximum rarity score of 20, which is awarded to records that were not on the UK species list as of November 2024 when we last updated our scoring system. All of these records have all been verified to Research Grade status on iNaturalist. N.B. some of the records here are based in the Channel Islands, which was included in the National Bioblitz but records from this region were not included when we created the rarity scoring system. It is possible these species are not as rare in the Channel Islands as they are in the UK.

### Non-native species

During National BioBlitz 2025 we had a specific focus on non-native species and encouraged people to record these species by doubling their rarity scores during the event. As a result, some of these records were also big scorers:

```{r best_non_native_recs, results='asis'}
#get photos
dir.create("images/best_non-native", showWarnings = FALSE)

for (i in 1:nrow(NatBioBlitz_iNat_NNS_best)) {
  
  url <- NatBioBlitz_iNat_NNS_best$observation_photos[[i]][1,]$photo.url
  
  #switch to medium image
  url <- gsub("square", "medium", url)
  
  if(!NatBioBlitz_iNat_NNS_best$license_code[[i]] %in% c("", "All Rights Reserved") | is.na(NatBioBlitz_iNat_NNS_best$license_code[[i]])){
  
  if (!is.na(url) && nzchar(url)) {
    filename <- paste0("images/best_non-native/native_", i, ".jpg")
    
    if(!filename %in% list.files("images/best_non-native")){
    download.file(url, destfile = filename, mode = "wb")
    NatBioBlitz_iNat_NNS_best$local_photo[i] <- filename  
    }
    
  } else {
    NatBioBlitz_iNat_NNS_best$local_photo[i] <- NA
  }
  }else{
    NatBioBlitz_iNat_NNS_best$local_photo[i] <- "License not available"
  }
}


# Loop through and print markdown for each record
for (i in 1:nrow(NatBioBlitz_iNat_NNS_best)) {
  rec <- NatBioBlitz_iNat_NNS_best[i, ]
  
  cat("\n\n")
  
  # Start minipage block (centered)
  cat("\\begin{center}\n")
  cat("\\begin{minipage}{0.9\\textwidth}\n")
  
  # Add image (sized down for PDF)
  if (rec$local_photo != "") {
    cat(sprintf("\\begin{center}\\includegraphics[height=6cm]{%s}\\end{center}\n\n", rec$local_photo))
  }
  
  # Print species and observation details
  if(!is.na(rec$taxon.preferred_common_name)){
  cat(sprintf("\\textbf{%s} \\textit{(%s)}\\\\\n", 
              rec$taxon.preferred_common_name, rec$taxon.name))
  }else{
    cat(sprintf("\\textbf{\\textit{%s}}\\\\\n", rec$taxon.name))
  }
  cat(sprintf("\\textbf{Rarity points:} %s\\\\\n", rec$Score))
  cat(sprintf("\\textbf{Location:} %s\\\\\n", rec$place_guess))
  
  if(!is.na(rec$user.name)){
    cat(sprintf("\\textbf{Observer:} %s\\\\\n", rec$user.name))
  }else{
    cat(sprintf("\\textbf{Observer:} %s\\\\\n", rec$user.login))
  }
  
  
  cat(sprintf("\\href{https://www.inaturalist.org/observations/%s}{View on iNaturalist}\\\\\n", rec$id))
  
    # Attribution if applicable
  if (!is.na(rec$observation_photos[[1]]$photo.attribution[1])) {
    cat(sprintf("\\textit{Photo licensed under CC-%s}\\\\\n", rec$observation_photos[[1]]$photo.attribution[1]))
  }

  # End minipage and add spacing
  cat("\\end{minipage}\n")
  cat("\\end{center}\n")
  cat("\\vspace{10mm}\n\n")
}

```


## Regional Summary

\begin{center}
\includegraphics[width=0.8\textwidth]{images/region map.png}
\end{center}

```{r region_breakdown}
#load region details
region_details <- read.csv("region names with inat codes.csv")

get_region <- function(place_ids, region_codes, region_names) {
  matched_index <- which(region_codes %in% place_ids)
  if (length(matched_index) == 0) return(NA)
  return(region_names[matched_index[1]])
}

# Map regions to each record
NatBioBlitz_iNat$region <- sapply(
  NatBioBlitz_iNat$place_ids,
  get_region,
  region_codes = region_details$iNat.code,
  region_names = region_details$Database.name
)

reg_recs <- table(NatBioBlitz_iNat$region)
reg_RG_recs <- table(subset(NatBioBlitz_iNat, quality_grade == "research")$region)

library(dplyr)
observer_counts <- NatBioBlitz_iNat %>%
  distinct(region, user.login) %>%
  count(region) 

observer_counts <- observer_counts$n[!is.na(observer_counts$region)]

reg_summary_tab <- data.frame("Region" = names(reg_recs),"Records" = as.vector(reg_recs), "Research Grade" = as.vector(reg_RG_recs), "Observers" = observer_counts)

library(kableExtra)

kable(format(reg_summary_tab, big.mark = ",", scientific = FALSE), col.names = gsub("\\."," ",names(reg_summary_tab))) %>%
  kable_styling(bootstrap_options = "striped")

no_reg_recs <- sum(is.na(NatBioBlitz_iNat$region))

library(english)

# Convert number to words with initial capital only
x <- as.character(english(no_reg_recs))
x <- gsub("-", " ", x)                    # Replace hyphen with space
x <- paste(toupper(substring(x, 1, 1)), substring(x, 2), sep = "")  # Capitalise first letter only


```

`r x` records could not be assigned to a region.

```{r reg_best_observers}
#find best


```



```{r regional-plot-sw}
# Placeholder for regional bar chart or map
```

\newpage

# Individual Feedback (optional)

Tailor info for a specific individual

## Example User Summary

- Total records: _X_
- Species recorded: _Y_
- Non-native species: _Z_

top species

```{r individual-feedback-example}
# Placeholder for individual bar/tile chart
```

\newpage

# Non-native Species Results

## Summary from the National BioBlitz

Overview of the non-native species found

```{r nn-summary}
# Placeholder for top non-native species chart
```

## Comparison with NBN Atlas Baseline

 - Which species were found in unusual places?
 - Which species weren't found in places they would be expected to be found?

```{r nn-comparison}
# Placeholder for before/after maps or bar charts
```

\newpage

# Species Highlights

- Rarest record
- Best photo submission
- Noteworthy range extension

```{r highlights}
# Placeholder for image collage or featured species panel
```


\newpage

# Project Partners and Support

Thank you to all our partners, volunteers, and funders.

ADD PARTNER LOGOS HERE

\newpage

# Next Steps: How to Stay Involved

## 1. Join or Start a Rock Pool Hub
- Get involved locally or start your own hub with our support.

## 2. Keep Recording on iNaturalist
- Every observation adds to the UK marine picture.

## 3. Seasonal Challenges
- Look out for themed events and mini-BioBlitzes.

## 4. Follow & Tag Us
- Stay connected via social media: [@therockpoolproject](https://www.instagram.com/therockpoolproject)

## 5. Volunteer or Collaborate
- Run events, support schools, or partner with us.

## 6. Sign Up for Updates
- Visit [www.therockpoolproject.co.uk](https://www.therockpoolproject.co.uk)

\newpage

# Appendices

## Species List

```{r species-table}
# Placeholder for table of all recorded species
```

## Methodology Notes

Details on data collection, validation, and tools used.

## Data Use Statement

Explanation of how participant and observation data are used, stored, and shared.
